# Statistical Analysis

In this section, we will dive into two R packages allowing us to leverage high-end distributed frameworks for statistical analysis: `h2o` and `sparklyr`.

## Platform `h2o`

`h2o` is a distributed platform for ML. You can download and run it on your local machine, but you could also deploy it on a server. The backend code is in Java and therefore runs super fast, it is particularly interesting at scale. See the tutorial `tutorial_4_h2o.R` and check out how it works (broadly) over here:

```{r}
#| eval: false

train_h2o_data <- as.h2o(...) # <- give some data

h2o.init() # start h2o local JAVA server

auto_ml <- h2o.automl(y=..., training_frame=train_h2o_data) # run auto ML

best_model <- h2o.get_best_model(auto_ml) # extract best model

prediction <- h2o.predict(best_model, test_h2o_data) # check out predictions
```

## Apache Spark

Apache Spark allows for distributed ML and analytics on data. You can work on Onyxia or in a local spark server to run the code in `tutorial_5_sparklyr.R`.

The process looks something along the following lines:

```{r}
#| eval: false

library(sparklyr) # load lib

sc <- spark_connect(master = "local") # connect to the server, local or not

flights_tbl <- copy_to(sc, nycflights13::flights, "flights") # copy data

model_delay <- flights_tbl %>%
  na.omit(arr_delay, distance) %>%
  ml_linear_regression(x=., formula=arr_delay~distance) # perform ML on the cluster
```

## Going further

You are very much encouraged to apply those methods to your own projects and see what works and what doesn't. As we've seen in the past chapters, there is no one solution and no solution fits all.

Something quite interesting would be to compare runtime when running usual R code versus those solutions. Happy exploration!